### 2025-03-03 scripts/Paper-analysis.Rmd
latest analysis

```
```{r}
ggplot(df_4_filter, aes(x="k=4", wilcoxon_stat, colour=p_value<0.05))+
  geom_point()
```

@@

@@

@@

```{r}
# k= 4
# 6-bp palindromes
palindromes = read.csv('~/Downloads/trieste/k4.txt', header=F)$V1
results.df = read.csv('../test_ptus_k4_scores.csv',header=T)
# Add PTUs as column names
colnames(results.df)[2:ncol(results.df)] = names(table(lr.no.duplicates.subset$PTU))
# Normalise scores by the number of plasmids within PTU?
n.plasmids = table(lr.no.duplicates.subset$PTU)
results.df.norm = results.df %>%
mutate(across(starts_with("PTU-"), ~ .x / n.plasmids[cur_column()], .names = "{.col}"))
# Plot all palindromes
results.df.long = results.df.norm %>%
pivot_longer(cols = starts_with("PTU"),
names_to = "score_type",
values_to = "score_avg")
results.df.norm.average = results.df.long %>% group_by(kmer) %>%
summarise(median=median(score_avg)) %>%
arrange(median)
# Take the top 1% depleted
results.df.norm.average.top.depleted = results.df.norm.average[which(results.df.norm.average$median< quantile(results.df.norm.average$median, probs = c(0.05))),] # not enough for 0.01
results.df.norm.average.top.depleted$palindrome = ifelse(results.df.norm.average.top.depleted$kmer %in% palindromes,
"palindrome",
"non-palindrome")
# Read in all RM targets
all_rm_targets = read.csv('../data/all_rebase_targets_250131_6bp_less_deambiguated.txt', header=F, sep=' ')
all_rm_targets_4bp =all_rm_targets$V1[all_rm_targets$V2==4]
results.df.norm.average.top.depleted$target = ifelse(results.df.norm.average.top.depleted$kmer %in%
all_rm_targets_4bp,
"target", "non-target")
table(results.df.norm.average.top.depleted$palindrome)
table(results.df.norm.average.top.depleted$target)
chisq.test(matrix(c(3, 10, 13, 230 ), nrow=2))
chisq.test(matrix(c(4, 9, 99, 144 ), nrow=2))

### And for k=5
results.df = read.csv('../test_ptus_k5_scores.csv',header=T)
# Add PTUs as column names
colnames(results.df)[2:ncol(results.df)] = names(table(lr.no.duplicates.subset$PTU))
# Normalise scores by the number of plasmids within PTU?
n.plasmids = table(lr.no.duplicates.subset$PTU)
results.df.norm = results.df %>%
mutate(across(starts_with("PTU-"), ~ .x / n.plasmids[cur_column()], .names = "{.col}"))
results.df.long = results.df.norm %>%
pivot_longer(cols = starts_with("PTU"),
names_to = "score_type",
values_to = "score_avg")
results.df.norm.average = results.df.long %>% group_by(kmer) %>%
summarise(median=median(score_avg)) %>%
arrange(median)
# Take the top 1% depleted
results.df.norm.average.top.depleted = results.df.norm.average[which(results.df.norm.average$median< quantile(results.df.norm.average$median, probs = c(0.01))),] # Top 5% since so few
all_rm_targets_5bp =all_rm_targets$V1[all_rm_targets$V2==5]
results.df.norm.average.top.depleted$target = ifelse(results.df.norm.average.top.depleted$kmer %in%
all_rm_targets_5bp,
"target", "non-target")
table(results.df.norm.average.top.depleted$target)
chisq.test(matrix(c(5, 6, 277, 736 ), nrow=2))

# k =6
# 6-bp palindromes
palindromes = read.csv('~/Downloads/trieste/k6.txt', header=F)$V1
results.df = read.csv('../test_ptus_k6_scores.csv',header=T)
# Add PTUs as column names
colnames(results.df)[2:ncol(results.df)] = names(table(lr.no.duplicates.subset$PTU))
# Normalise scores by the number of plasmids within PTU?
n.plasmids = table(lr.no.duplicates.subset$PTU)
results.df.norm = results.df %>%
mutate(across(starts_with("PTU-"), ~ .x / n.plasmids[cur_column()], .names = "{.col}"))
# Plot all palindromes
results.df.long = results.df.norm %>%
pivot_longer(cols = starts_with("PTU"),
names_to = "score_type",
values_to = "score_avg")
results.df.norm.average = results.df.long %>% group_by(kmer) %>%
summarise(median=median(score_avg)) %>%
arrange(median)
# Take the top 1% depleted
results.df.norm.average.top.depleted = results.df.norm.average[which(results.df.norm.average$median< quantile(results.df.norm.average$median, probs = c(0.01))),]
results.df.norm.average.top.depleted$palindrome = ifelse(results.df.norm.average.top.depleted$kmer %in% palindromes,
"palindrome",
"non-palindrome")
# Read in all RM targets
all_rm_targets = read.csv('../data/all_rebase_targets_250131_6bp_less_deambiguated.txt', header=F, sep=' ')
all_rm_targets_6bp =all_rm_targets$V1[all_rm_targets$V2==6]
results.df.norm.average.top.depleted$target = ifelse(results.df.norm.average.top.depleted$kmer %in%
all_rm_targets_6bp,
"target", "non-target")
table(results.df.norm.average.top.depleted$palindrome)
table(results.df.norm.average.top.depleted$target)
chisq.test(matrix(c(7, 34, 865-7, 4096-865-34 ), nrow=2))
chisq.test(matrix(c(4, 37, 60, 4055 ), nrow=2))

>>>>

@@

Repeated this analysis with one gene representative per family:

```{r}
# Read in results for RM targets
df_4 = read.csv('../test_one_per_family_within_range_targets_k4.csv')
df_5 = read.csv('../test_one_per_family_within_range_targets_k5.csv')
df_6 = read.csv('../test_one_per_family_within_range_targets_k6.csv')

# Convert discrepancies to numeric (presence of None has messed this up)
df_4$discrepancy_selected = as.numeric(df_4$discrepancy_selected)
df_4$discrepancy_selected[is.na(df_4$discrepancy_selected)] = 0
df_5$discrepancy_selected = as.numeric(df_5$discrepancy_selected)
df_5$discrepancy_selected[is.na(df_5$discrepancy_selected)] = 0
df_6$discrepancy_selected = as.numeric(df_6$discrepancy_selected)
df_6$discrepancy_selected[is.na(df_6$discrepancy_selected)] = 0

# Median of selected k-mers
df = t(data.frame(c(median(df_4[which(df_4$n_selected_kmers!=0),"discrepancy_selected"]), 
        wilcox.test(df_4[which(df_4$n_selected_kmers!=0),"discrepancy_selected"])$statistic,
        wilcox.test(df_4[which(df_4$n_selected_kmers!=0),"discrepancy_selected"])$p.value),
        c(median(df_5[which(df_5$n_selected_kmers!=0),"discrepancy_selected"]), 
        wilcox.test(df_5[which(df_5$n_selected_kmers!=0),"discrepancy_selected"])$statistic,
        wilcox.test(df_5[which(df_5$n_selected_kmers!=0),"discrepancy_selected"])$p.value),
        c(median(df_6[which(df_6$n_selected_kmers!=0),"discrepancy_selected"]), 
        wilcox.test(df_6[which(df_6$n_selected_kmers!=0),"discrepancy_selected"])$statistic,
        wilcox.test(df_6[which(df_6$n_selected_kmers!=0),"discrepancy_selected"])$p.value)))
colnames(df) = c("median", "statistic", "p.value")
print(df)

### All hard shell vs. accessory, one representative per gene family

We subsampled to one representative per gene family and the same number of hard shell and accessory genes, whichever was smaller (PTU-Bac37 was the only one that had more hard shell than accessory genes). 
```{r}
palindromes_k4 = tolower(read.csv('~/Downloads/trieste/k4.txt', header=F)$V1)
palindromes_k6 = tolower(read.csv('~/Downloads/trieste/k6.txt', header=F)$V1)

core_k6 = read.csv('~/Downloads/trieste/one_per_family/all_core_1745k_k6.csv')
accessory_k6 = read.csv('~/Downloads/trieste/one_per_family/all_accessory_1745k_k6.csv')
merge_k6 = merge(core_k6, accessory_k6, by="word")
merge_k6$score.diff = merge_k6$score.x-merge_k6$score.y
merge_k6$count.diff = merge_k6$count.x-merge_k6$count.y

merge_k6$palindrome = ifelse(merge_k6$word %in% palindromes_k6,
                             "palindrome", "other")
median(merge_k6[which(merge_k6$palindrome=="palindrome"),"score.diff"])
wilcox.test(merge_k6[which(merge_k6$palindrome=="palindrome"),"score.diff"])
sum(merge_k6$count.x[which(merge_k6$palindrome=="palindrome")])/sum(merge_k6$count.x) * 1000
sum(merge_k6$count.y[which(merge_k6$palindrome=="palindrome")])/sum(merge_k6$count.y) * 1000

# And for k=4

core_k4 = read.csv('~/Downloads/trieste/one_per_family/all_core_1745k_k4.csv')
accessory_k4 = read.csv('~/Downloads/trieste/one_per_family/all_accessory_1745k_k4.csv')
merge_k4 = merge(core_k4, accessory_k4, by="word")
merge_k4$score.diff = merge_k4$score.x-merge_k4$score.y
merge_k4$count.diff = merge_k4$count.x-merge_k4$count.y

merge_k4$palindrome = ifelse(merge_k4$word %in% palindromes_k4,
                             "palindrome", "other")
median(merge_k4[which(merge_k4$palindrome=="palindrome"),"score.diff"])
wilcox.test(merge_k4[which(merge_k4$palindrome=="palindrome"),"score.diff"])

# Sum palindromes
sum(merge_k4$count.x[which(merge_k4$palindrome=="palindrome")])/sum(merge_k4$count.x) * 1000
sum(merge_k4$count.y[which(merge_k4$palindrome=="palindrome")])/sum(merge_k4$count.y) * 1000

@@

```
```{r}
# Sliding window for GC content
p.lines.GC = ggplot(d.reduced.median.plot, aes(x, GC/5000 * 100))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome depletion per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.GC, file='../results/GC-content.pdf',
       width=10, height=6)

# And for palindrome density (not expected - actual values)
p.lines.density = ggplot(d.reduced.median.plot, aes(x, median.original*1000))+
  geom_hline(yintercept = 0, linetype='dashed')+
  theme_bw()+
  geom_path(aes(group=PTU, alpha=normalized_n))+
  facet_wrap(~PTU, scales="free_x", nrow=3)+
  ylab("Palindrome density per kb (5kb sliding window)")+
  labs(alpha="% of plasmids within PTU")+
  theme(panel.grid=element_blank())+  
  scale_x_continuous(labels = function(x) x / 1000) +
  xlab("Position on plasmid (kb)")+
  theme(legend.position = c(0.8, 0.2))
ggsave(p.lines.density, file='../results/6mer-palindrome-density.pdf',
       width=10, height=6)

@@

```{r}
# Normalise
PTU.counts = table(plasmid.table$PTU)
n.plasmids = PTU.counts[leading.region.PTUs]
# k= 4
# 6-bp palindromes
palindromes = read.csv('~/Downloads/trieste/k4.txt', header=F)$V1
results.df = read.csv('../test_ptus_k4_scores.csv',header=T)
# Add PTUs as column names
colnames(results.df)[2:ncol(results.df)] = leading.region.PTUs
results.df.norm = results.df %>%
mutate(across(starts_with("PTU-"), ~ .x / n.plasmids[cur_column()], .names = "{.col}"))
# Plot all palindromes
results.df.long = results.df.norm %>%
pivot_longer(cols = starts_with("PTU"),
names_to = "score_type",
values_to = "score_avg")
results.df.norm.average = results.df.long %>% group_by(kmer) %>%
summarise(median=median(score_avg)) %>%
arrange(median)
# Take the top 1% depleted
results.df.norm.average.top.depleted = results.df.norm.average[which(results.df.norm.average$median< quantile(results.df.norm.average$median, probs = c(0.05))),] # not enough for 0.01
results.df.norm.average.top.depleted$palindrome = ifelse(results.df.norm.average.top.depleted$kmer %in% palindromes,
"palindrome",
"non-palindrome")
# Read in all RM targets
all_rm_targets = read.csv('../data/rebase_all_4_6bp_targets_250131.csv', header=T)
all_rm_targets_4bp =all_rm_targets$target[all_rm_targets$length==4]
results.df.norm.average.top.depleted$target = ifelse(results.df.norm.average.top.depleted$kmer %in%
all_rm_targets_4bp,
"target", "non-target")
table(results.df.norm.average.top.depleted$palindrome)
table(results.df.norm.average.top.depleted$target)
# palindromes
chisq.test(matrix(c(2, 11, 13, 230 ), nrow=2))
# Targets
chisq.test(matrix(c(3, 10, 100, 143 ), nrow=2))

### And for k=5
results.df = read.csv('../test_ptus_k5_scores.csv',header=T)
# Add PTUs as column names
colnames(results.df)[2:ncol(results.df)] = leading.region.PTUs
results.df.norm = results.df %>%
mutate(across(starts_with("PTU-"), ~ .x / n.plasmids[cur_column()], .names = "{.col}"))
results.df.long = results.df.norm %>%
pivot_longer(cols = starts_with("PTU"),
names_to = "score_type",
values_to = "score_avg")
results.df.norm.average = results.df.long %>% group_by(kmer) %>%
summarise(median=median(score_avg)) %>%
arrange(median)
# Take the top 1% depleted
results.df.norm.average.top.depleted = results.df.norm.average[which(results.df.norm.average$median< quantile(results.df.norm.average$median, probs = c(0.01))),] 
all_rm_targets_5bp =all_rm_targets$target[all_rm_targets$length==5]
results.df.norm.average.top.depleted$target = ifelse(results.df.norm.average.top.depleted$kmer %in%
all_rm_targets_5bp,
"target", "non-target")
table(results.df.norm.average.top.depleted$target)
chisq.test(matrix(c(5, 11, 277, 731 ), nrow=2))

# k =6
# 6-bp palindromes
palindromes = read.csv('~/Downloads/trieste/k6.txt', header=F)$V1
results.df = read.csv('../test_ptus_k6_scores.csv',header=T)
# Add PTUs as column names
colnames(results.df)[2:ncol(results.df)] = leading.region.PTUs
results.df.norm = results.df %>%
mutate(across(starts_with("PTU-"), ~ .x / n.plasmids[cur_column()], .names = "{.col}"))
# Switch to long format
results.df.long = results.df.norm %>%
pivot_longer(cols = starts_with("PTU"),
names_to = "score_type",
values_to = "score_avg")

results.df.norm.average = results.df.long %>% group_by(kmer) %>%
summarise(median=median(score_avg)) %>%
arrange(median)
# Take the top 1% depleted
results.df.norm.average.top.depleted = results.df.norm.average[which(results.df.norm.average$median< quantile(results.df.norm.average$median, probs = c(0.01))),]
results.df.norm.average.top.depleted$palindrome = ifelse(results.df.norm.average.top.depleted$kmer %in% palindromes,
"palindrome",
"non-palindrome")
# Read in all RM targets
all_rm_targets = read.csv('../data/rebase_all_4_6bp_targets_250131.csv', header=F, sep=' ')
all_rm_targets_6bp =all_rm_targets$target[all_rm_targets$length==6]
results.df.norm.average.top.depleted$target = ifelse(results.df.norm.average.top.depleted$kmer %in%
all_rm_targets_6bp,
"target", "non-target")
table(results.df.norm.average.top.depleted$palindrome)
table(results.df.norm.average.top.depleted$target)
chisq.test(matrix(c(7, 34, 865-7, 4096-865-34 ), nrow=2))
chisq.test(matrix(c(4, 37, 60, 4055 ), nrow=2))
